\proglang{Stan}: A Probabilistic Programming Language

### 2013-12-25 src/docs/papers/stan-language/article.tex
updated first half of paper

@@

%% an abstract and keywords
\Abstract{ \proglang{Stan} is a probabilistic programming language for
  specifying statistical models. A \proglang{Stan} program
  imperatively defines a log probability function over parameters
  conditioned on specified data and constants.

  Full Bayesian inference is supported through Markov chain Monte
  Carlo (MCMC) methods such as the No-U-Turn sampler (NUTS), an
  adaptive form of Hamiltonian Monte Carlo (HMC) sampling.  Penalized
  maximum likelihood estimates are calculated using optimization
  methods such as BFGS.

@@

  Stan can be viewed as a package for performing full Bayesian
  inference for continuous-variable models, and as a platform for
  computing log-densities and their gradients and Hessians, which can
  be used in alternative algorithms such as variational Bayes,
  expectation propagation, and marginal inference using approximate
  integration.  To this end, Stan is set up so that the densities,
  gradients, and Hessians, along with intermediate quantities of the
  algorithm such as acceptance probabilities, are easily accessible.

  % \proglang{Stan} can be called from the command line, through
  % \proglang{R} using the \pkg{RStan} package, or through
  % \proglang{Python} using the \pkg{PyStan} package.  All three
  % interfaces support sampling or optimization-based inference and
  % analysis.
}

@@

We did not set out to build \proglang{Stan} as it currently exists.%
%
\footnote{\proglang{Stan}'s home page, \url{http://mc-stan.org/},
  links to the Stan manual, example models including translations of
  the BUGS examples and the models from \citep{GelmanHill:2007},
  getting started instructions and full documentation for Stan's
  interfaces for the command line (CmdStan), Python (PyStan), and
  R (Rstan), and the source code repositories and issue trackers.}

Our original goal was to apply full Bayesian inference to the sort of multilevel
generalized linear models discussed in Part~II of
\citep{GelmanHill:2007}, which are structured with grouped and
interacted predictors at multiple levels, hierarchical covariance
priors, nonconjugate coefficient priors, latent effects as in
item-response models, and varying output link functions and
distributions.

These models turned out to be a challenge for
existing general-purpose inference software.  A direct encoding in
\proglang{BUGS} \citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
\proglang{JAGS} \citep{Plummer:2003} can grind these tools to a halt.
We began by adding custom vectorized logistic regressions to \proglang{JAGS}
using \proglang{C++} to overcome the cost of interpretation.  Although
this is much faster than looping in \proglang{JAGS}, it quickly became
clear that the root of the problem was the slow convergence of
conditional sampling when parameters were highly correlated 
in the posterior, as for time-series models and hierarchical models 
with interacted predictors.  We finally realized we needed a better sampler, 
not a more efficient implementation.

@@

We were hearing more about HMC (Hamiltonian Monte Carlo), which
appeared promising but was also problematic in that the Hamiltonian
dynamics simulation requires the gradient of the log posterior.
Although possible, computing these gradients by hand on a model-by-model basis
is very tedious and error prone.  That is when we discovered
reverse-mode algorithmic differentiation, which, given a
templated \proglang{C++} function for the log posterior,
automatically computes a proper analytic gradient up to machine
precision accuracy in only a few multiples of the cost to evaluate the
log probability function itself.  We explored existing algorithmic
differentiation packages with open licenses such as \pkg{RAD}
\citep{Gay:2005} and its repackaging in the Sacado module of the
\pkg{Trilinos} toolkit, and the \pkg{CppAD} package of the COIN-OR
toolkit (see \citep{BellBurke:2008}).  Both packages, however, supported
few special functions (e.g., probability functions, log gamma,
inverse logit) or linear algebra operations (e.g., Cholesky
decompositions, matrix division), and neither are easily or modularly
extensible.

Consequently we built our own reverse-mode algorithmic differentiation package.
At that point, we ran into the problem that we could not just plug in
the probability functions from a package like \pkg{Boost} because they
weren't templated generally enough across all the arguments.  Rather
than pay the price of promoting floating point values to algorithmic
differentiation variables, we wrote our own fully templated
probability functions and other special functions.

Next, we integrated the \proglang{C++} package \pkg{Eigen} for matrix
operations and linear algebra functions.  \pkg{Eigen} makes extensive
use of expression templates for lazy evaluation and the CRTP
(curiously recurring template pattern) to implement concepts without
virtual function calls; \cite{VandevoordeJosuttis:2002} provide a 
complete description of template metaprogramming techniques. 
Unfortunately, we
ran into the same problem with Eigen as with the existing probability
libraries---it doesn't support mixed operations of algorithmic
differentiation variables and primitives like \code{double}.  Although
we initially began by promoting floating-point vectors to algorithmic
differentiation variables, we later completely rewrote our matrix
library to efficiently compute derivatives using rules described by
\cite{Giles:2008}.

At this point, we could fit models coded directly in \proglang{C++} on
top of the pre-release versions of the \proglang{Stan} API
(application programming interface).  Seeing how well this all worked,
we set our sights on the generality and ease of use of
\proglang{BUGS} and designed a modeling language in which
statisticians could write their models in familiar notation that could
be transformed to efficient \proglang{C++} code before being compiled into
an efficient executable program.  This paper is primarily about this
language and how it specifies models.

@@

Even with the prototype compiler generating models, we still faced a
major hurdle for ease of use. The efficiency of HMC is very sensitive
to two tuning parameters, the discretization interval (i.e., step
size) and the total simulation time (i.e., number of steps).  The interval
size parameter could be tuned during warm-up based on Metropolis
rejection rates, but the number of steps proved difficult to tune without
sacrificing the detailed balance of the sampler.  This led to the
development of the NUTS (No U-Turn) sampler
\citep{HoffmanGelman:2011}.  Roughly speaking, NUTS builds a tree
of possible samples by randomly simulating both forwards and backwards
in time until the combined trajectory turns back on itself.  Once the trajectory
has terminated, a new sample is drawn from the tree.

We thought we were home free at this point, but when we measured the
speed of some \proglang{BUGS} examples versus \proglang{Stan} we were
very disappointed.  \proglang{BUGS}'s very first example model, Rats,
ran more than an order of magnitude faster in \proglang{JAGS} than in
\proglang{Stan}.  Rats is a tough test case because the conjugate
priors and lack of posterior correlations make it an ideal candidate
for efficient Gibbs sampling.

Realizing that we were doing redundant calculations, we wrote a
vectorized form of the normal distribution for multiple variates with
the same mean and scale, which sped things up a bit. At the same time,
we introduced some simple template metaprograms to remove the
calculation of constant terms in the log probability.  Performance was
lacking until we finally figured out how to both eliminate redundant 
calculations and partially evaluate the gradients using
a combination of expression templates and metaprogramming.

When we attempted to fit a time-series model, we found that
normalizing the data to unit sample mean and variance sped up the fits
by an order of magnitude.  In hindsight this isn't surprising, as the
performance of the numerical simulation in HMC scales with the
variation of the parameter scales; fortunately HMC also provides a
choice of metric, or mass matrix, to compensate for variations in the
log posterior.  In \proglang{Stan} 1.0 we introduced an adaptive
diagonal metric into our HMC implementations that allowed the
parameter scales to be normalized automatically, with the option of an
adaptive dense metric and a full automated standardization of the
posterior in \proglang{Stan} 2.0.

Because the dense metric performs only a global standardization, the
performance of \proglang{Stan} in models where the relative parameter
scales varies by location, such as hierarchical and latent models, can
suffer.  Riemannian manifold Hamiltonian Monte Carlo (RHMC) introduces
a location-dependent metric that can overcome these final hurdles
\citep{GirolamiCalderhead:2011}; Stan has a prototype implementation
of RHMC based on the metric of \citep{Betancourt:2012}, using a
generalization of NUTS to Riemann manifolds \citep{Betancourt:2013}.

@@

This section describes the use of \proglang{Stan} from the command
line for estimating (i.e., sampling from) the posterior of a Bayesian
model.

@@

Consider estimating the chance of success parameter for a Bernoulli
distribution based on a sequence of observed binary outcomes.  
Figure~\ref{bernoulli-model.fig} provides an implementation of such a
model in \proglang{Stan}.
%
\begin{figure}
\begin{Code}
data { 
  int<lower=0> N; 
  int<lower=0,upper=1> y[N];
} 
parameters {
  real<lower=0,upper=1> theta;
} 
model {
  theta ~ beta(1,1);     // prior
  y ~ bernoulli(theta);  // likelihood
}
\end{Code}
\caption{Model for estimating a Bernoulli parameter.}\label{bernoulli-model.fig}
\end{figure}
%
The model treats the observed binary data, \code{y[1],...,y[N]}, as
independent and identically distributed, with success probability
\code{theta}.  The vectorized likelihood statement can also be coded
using a loop as in BUGS, although it will run more slowly than the
vectorized form:
%
\begin{Code}
for (n in 1:N)
  y[n] ~ bernoulli(theta);
\end{Code}
%
A \code{beta(1,1)} (i.e., uniform) prior is placed on \code{theta},
although there is no special behavior for conjugate priors in
\proglang{Stan}.  The prior could be dropped from the model altogether
because parameters start with uniform distributions on their support,
here constrained to be between 0 and 1 in the parameter declaration
for \code{theta}.  This model is available in the \proglang{Stan}
source distribution in
\code{src/models/basic\_estimators/bernoulli.stan}.

@@

Data for running \proglang{Stan} from the command line can be included
in \proglang{R} dump format.%
%
\footnote{A JSON interface for structured data input and output is
  currently under development.}
%
For example, 10 observations for the
model in Figure~\ref{bernoulli-model.fig} could be encoded as
%
\begin{Code}
N <- 10
y <- c(0,1,0,0,0,0,0,0,0,1)
\end{Code}
%
This defines the contents of two variables, an integer \code{N} and a
10-element integer array \code{y}.  In \pkg{RStan} and \pkg{PyStan},
data can also be passed directly through memory without the need to
read or write to a file.  This data file is provided with the
\proglang{Stan} distrbution in file
\code{src/models/basic\_estimators/bernoulli.R.stan}.

@@

\subsection{Running the model}

\subsubsection{Command to run the model}

The executable can be run with default options by specifying a path to
the data file.  The first command in the following example changes the
current directory to that containing the model, which is where the
data resides and where the executable is built.  From there, the path
to the data is just the file name \code{bernoulli.data.R}.
%
\begin{CodeChunk}
{\small
\begin{CodeInput}
> cd $stan/src/models/basic_estimators
> ./bernoulli --data=bernoulli.data.R --samples=samp1.csv --chain_id=1 --seed=7386
\end{CodeInput}
}
\end{CodeChunk}
%
For Windows, the \code{./} before the command should be removed.  This
invocation specifies the executable program, the data file, the file
to which the samples are written, the identity of the chain, and a
seed for the random number generator.

\subsubsection{Terminal output}

The output is as follows, starting with a summary of the command-line
options used, including defaults;  these are also written into the 
samples file as comments.
%
\begin{Code}
STAN SAMPLING COMMAND
data = bernoulli.data.R
init = random initialization
init tries = 1
samples = samp1.csv
append_samples = 0
save_warmup = 0
seed = 7386 (user specified)
chain_id = 1 (user specified)
iter = 2000
warmup = 1000
thin = 1
equal_step_sizes = 0
nondiag_mass = 0
leapfrog_steps = -1
max_treedepth = 10
epsilon = -1
epsilon_pm = 0
delta = 0.5
gamma = 0.05
Iteration:    1 / 2000 [  0%]  (Warmup)
Iteration:   10 / 2000 [  0%]  (Warmup)
...
Iteration: 1990 / 2000 [ 99%]  (Sampling)
Iteration: 2000 / 2000 [100%]  (Sampling)

# Elapsed Time: 0.011269 seconds (Warm-up)
#               0.014105 seconds (Sampling)
#               0.025374 seconds (Total)
\end{Code}
%
A description of all parameters is available with the \code{{-}{-}help}
option; they provide control over input, output, and the configuration
of the sampler.

@@

The output CSV file starts with a summary of the parameters for the
run.  
%
\begin{Code}
# Samples Generated by Stan
#
# stan_version_major=1
# stan_version_minor=3
# stan_version_patch=0
# model=bernoulli_model
# data=bernoulli.data.R
# init=random initialization
# append_samples=0
# save_warmup=0
# seed=7386
# chain_id=1
# iter=2000
# warmup=1000
# thin=1
# nondiag_mass=0
# equal_step_sizes=0
# leapfrog_steps=-1
# max_treedepth=10
# epsilon=-1
# epsilon_pm=0
# delta=0.5
# gamma=0.05
# algorithm=NUTS with a diagonal Euclidean metric
\end{Code}
%
It continues with a header indicating the columns of the output.
%
\begin{Code}
lp__,accept_stat__,stepsize__,treedepth__,theta
\end{Code}
%
The values for \code{lp\_\_} indicate the log probability (up to an
additive constant).  The column headed by \code{accept\_stat\_\_}
provides the Metropolis acceptance statistic for each iteration.%
%
\footnote{Acceptance is the usual notion for a Metropolis-Hastings
  sampling such as HMC.  For NUTS, the acceptance statistic is 
  defined as the average acceptance probabilities of all possible
  samples in the proposed tree.}
%
The column \code{stepsize\_\_} indicates the step size
(i.e., time interval) of the simulated trajectory, while the column 
\code{treedepth\_\_} gives the tree depth for NUTS, defined as the 
log base 2 of the total number of steps in the trajectory.
The rest of the header will be the names of parameters; in this
example, \code{theta} is the only parameter.

Next, the results of adaptation are printed.
%
\begin{Code}
# Adaptation terminated
# Step size = 2.10898
# Diagonal elements of inverse mass matrix:
# 0.465584
\end{Code}
%
By default, the execution above uses a diagonal mass matrix, which
with one parameter is a single element.

The rest of the file contains samples, one per line, matching the
header; here the parameter \code{theta} is the final value printed on
each line, and each line corresponds to a sample.  The warmup samples
are not included by default, but may be included with the appropriate
command-line invocation of the executable. The file ends with comments reporting the elapsed time.
%
\begin{Code}
-7.04574,0.189209,2.10898,0,0.353589
-7.04574,0.00162111,2.10898,0,0.353589
-6.77448,1,2.10898,0,0.222005
...
-7.89094,0.346733,2.10898,0,0.46154
-6.93412,1,2.10898,0,0.330879
-6.93412,0.115548,2.10898,0,0.330879

# Elapsed Time: 0.011269 seconds (Warm-up)
#               0.014105 seconds (Sampling)
#               0.025374 seconds (Total)
\end{Code}
%

\subsection{Output analysis}

Before performing output analysis, we recommend generating multiple
independent chains in order to more effectively monitor convergence.
Three more chains of samples can be created as follows.
%
\begin{CodeChunk}
{\small
\begin{CodeInput}
> ./bernoulli --data=bernoulli.data.R --samples=samp2.csv --chain_id=2 --seed=7386
> ./bernoulli --data=bernoulli.data.R --samples=samp3.csv --chain_id=3 --seed=7386
> ./bernoulli --data=bernoulli.data.R --samples=samp4.csv --chain_id=4 --seed=7386
\end{CodeInput}
}
\end{CodeChunk}
%
The chains can be safely run in parallel under different processes;
details of parallel execution depend on the operating system and the shell or
terminal program. Note that, although the same seed is used for each chain, the
random numbers will in fact be independent as the chain identifier
is used to skip the pseudorandom number generator ahead.  See
Section~\ref{rng.section} for more information.

\proglang{Stan} supplies a command-line program \code{bin/print} to
summarize the output of one or more MCMC chains.  Given a directory
containing output from sampling, it is invoked as
%
\begin{CodeChunk}
\begin{CodeInput}
> ls samp*.csv
\end{CodeInput}
\begin{CodeOutput}
samp1.csv	samp2.csv	samp3.csv	samp4.csv
\end{CodeOutput}
\begin{CodeInput}
> ../../../bin/print samp*.csv
\end{CodeInput}

>>>>

\renewcommand{\textfraction}{0.07}
\renewcommand{\topfraction}{0.9}	% max fraction of floats at top

@@

%% an abstract and keywords
\Abstract{ \proglang{Stan} is a probabilistic programming language for
  specifying statistical models. A \proglang{Stan} program
  imperatively defines a log probability function over parameters
  conditioned on specified data and constants.  As of version 2.1,
  \proglang{Stan} provides full Bayesian inference for continuous-variable
  models through Markov chain Monte Carlo (MCMC) methods such as
  the No-U-Turn sampler (NUTS), an adaptive form of Hamiltonian Monte
  Carlo (HMC) sampling.  Penalized maximum likelihood estimates are
  calculated using optimization methods such as BFGS.

@@

  \proglang{Stan} is also a platform for computing log-densities and
  their gradients and Hessians, which can be used in alternative
  algorithms such as variational Bayes, expectation propagation, and
  marginal inference using approximate integration.  To this end, Stan
  is set up so that the densities, gradients, and Hessians, along with
  intermediate quantities of the algorithm such as acceptance
  probabilities, are easily accessible.

  \proglang{Stan} can be called from the command line, through
  \proglang{R} using the \pkg{RStan} package, or through
  \proglang{Python} using the \pkg{PyStan} package.  All three
  interfaces support sampling or optimization-based inference and
  analysis, and \pkg{RStan} and \pkg{PyStan} also provide access to
  log probabilities, gradients, Hessians, and I/O transforms.
}

@@

We did not set out to build \proglang{Stan} as it currently exists.%
%
\footnote{\proglang{Stan}'s home page, \url{http://mc-stan.org/},
  links to the Stan manual, example models including translations of
  the BUGS examples and the models from \citep{GelmanHill:2007},
  getting started instructions and full documentation for Stan's
  interfaces for the command line shell (\pkg{CmdStan}), \proglang{Python}
  (\pkg{PyStan}), and \proglang{R} (\pkg{RStan}), and the source code repositories
  and issue trackers.}
%
Our original goal was to apply full Bayesian inference to the sort of multilevel
generalized linear models discussed in Part~II of
\citep{GelmanHill:2007}, which are structured with grouped and
interacted predictors at multiple levels, hierarchical covariance
priors, nonconjugate coefficient priors, latent effects as in
item-response models, and varying output link functions and
distributions.

These models turned out to be a challenge for
existing general-purpose inference software.  A direct encoding in
\proglang{BUGS} \citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
\proglang{JAGS} \citep{Plummer:2003} can grind these tools to a halt.
We began by adding custom vectorized logistic regressions to \proglang{JAGS}
using \proglang{C++} to overcome the cost of interpretation.  Although
this is much faster than looping in \proglang{JAGS}, it quickly became
clear that the root of the problem was the slow convergence of
conditional sampling when parameters were highly correlated 
in the posterior, as for time-series models and hierarchical models 
with interacted predictors.  We finally realized we needed a better sampler, 
not a more efficient implementation of Gibbs sampling.

@@

We were hearing more about Hamiltonian Monte Carlo (HMC) sampling,
which appeared promising but was also problematic in that the
Hamiltonian dynamics simulation requires the gradient of the log
posterior.  Although not difficult in theory for most functions,
computing these gradients by hand on a model-by-model basis is very
tedious and error prone.  That is when we discovered reverse-mode
algorithmic differentiation, which, given a templated \proglang{C++}
function for the log posterior, automatically computes a proper
analytic gradient up to machine precision accuracy in only a few
multiples of the cost to evaluate the log probability function itself.
We explored existing algorithmic differentiation packages with open
licenses such as \pkg{RAD} \citep{Gay:2005} and its repackaging in the
Sacado module of the \pkg{Trilinos} toolkit, and the \pkg{CppAD}
package of the COIN-OR toolkit \citep{BellBurke:2008}.  Both
packages, however, supported few special functions (e.g., probability
functions, log gamma, inverse logit) or linear algebra operations
(e.g., Cholesky decompositions, matrix division), and neither are
easily or modularly extensible.

Consequently we built our own reverse-mode algorithmic differentiation package.
At that point, we ran into the problem that we could not just plug in
the probability functions from a package like \pkg{Boost} because they
weren't templated generally enough across all arguments.  Rather
than pay the price of promoting floating point values to algorithmic
differentiation variables, we wrote our own fully templated
probability functions and other special functions.

Next, we integrated the \proglang{C++} package \pkg{Eigen} for matrix
operations and linear algebra functions.  \pkg{Eigen} makes extensive
use of expression templates for lazy evaluation and the curiously
recurring template pattern (CRTP) to implement concepts without
virtual function calls; \cite{VandevoordeJosuttis:2002} provide a
complete description of template metaprogramming techniques.
Unfortunately, we ran into the same problem with Eigen as with the
existing probability libraries---it doesn't support mixed operations
of algorithmic differentiation variables and primitives like
\code{double}.  Although we initially began by promoting
floating-point vectors to algorithmic differentiation variables, we
later completely rewrote mixed-mode operations like multiplying a data
matrix by a parameter vector.  To compute derivatives of matrix
operations such as division and inverse, we implemented specialized
derivatives using the rules described by \cite{Giles:2008}.

At this point, we could fit models coded directly in \proglang{C++} on
top of the pre-release versions of the \proglang{Stan} API
(application programming interface).  Seeing how well this all worked,
we set our sights on the generality and ease of use of \proglang{BUGS}
and designed a modeling language in which statisticians could write
their models in familiar notation that could be transformed to
efficient \proglang{C++} code before being compiled into an efficient
executable program.  Although we started with the target of specifying
directed graphical models, our first implementation turned out to be a
more general imperative form of specifying log probability functions;
\proglang{BUGS} models could be translated line for line, but the
\proglang{Stan} language also supports much richer imperative
constructs like conditionals, while loops, and local variables.
This paper is primarily about the \proglang{Stan} language and how
it can be used to specify statistical models.

@@

Even with the prototype compiler generating models, we still faced a
major hurdle for ease of use. The efficiency of HMC is very sensitive
to two tuning parameters, the discretization interval (i.e., step
size) and the total simulation time (i.e., number of steps).  The
interval size parameter could be tuned during warm-up based on
Metropolis rejection rates, but the number of steps proved difficult
to tune without sacrificing the detailed balance of the sampler.  This
led to the development of the No U-Turn (NUTS) sampler
\citep{HoffmanGelman:2011}.  Roughly speaking, NUTS builds a tree of
possible samples by randomly simulating Hamiltonian dynamics both
forwards and backwards in time until the combined trajectory turns
back on itself.  Once the trajectory has terminated, a new sample is
drawn from the tree.

We thought we were home free at this point, but when we measured the
speed of some \proglang{BUGS} examples versus \proglang{Stan} we were
very disappointed.  \proglang{BUGS}'s very first example model, Rats,
ran more than an order of magnitude faster in \proglang{JAGS} than in
\proglang{Stan}.  Rats is a tough test case because the conjugate
priors and lack of posterior correlations make it an ideal candidate
for efficient Gibbs sampling.  Realizing that we were doing redundant
calculations, we wrote a vectorized form of the normal distribution
for multiple variates with the same mean and scale, which sped things
up a bit. At the same time, we introduced some simple template
metaprograms to remove the calculation of constant terms in the log
probability.  Performance was lacking until we finally figured out how
to both eliminate redundant calculations and partially evaluate the
gradients using a combination of expression templates and
metaprogramming.

When we attempted to fit a time-series model, we found that
normalizing the data to unit sample mean and variance sped up the fits
by an order of magnitude.  In hindsight this isn't surprising, as the
performance of the numerical simulation in HMC scales with the
variation of the parameter scales; fortunately HMC also provides a
choice of metric, in our case a Euclidean metric with a constant mass
matrix, to compensate for variations in the log posterior.  In
\proglang{Stan} 1.0 we introduced an adaptive diagonal metric into our
HMC implementations that allowed the parameter scales to be normalized
automatically, with the option of an adaptive dense metric and a full
automated standardization of the posterior in \proglang{Stan} 2.0.

Because the dense metric performs only a global standardization, the
performance of \proglang{Stan} in models where the relative parameter
scales varies by location, such as hierarchical and latent models, can
suffer.  Riemannian manifold Hamiltonian Monte Carlo (RHMC) introduces
a location-dependent metric that can overcome these final hurdles
\citep{GirolamiCalderhead:2011}; Stan has a prototype implementation
of RHMC based on the SoftAbs metric of \citep{Betancourt:2012}, using
a generalization of NUTS to Riemannian manifolds \citep{Betancourt:2013}.

@@

This section describes the use of \proglang{Stan} from the command
line for estimating a Bayesian model using both MCMC sampling for full
Bayesian inference and optimization to provide a point estimate at the
posterior mode.

@@

Consider estimating the chance of success parameter for a Bernoulli
distribution based on a sequence of observed binary outcomes.  
Figure~\ref{bernoulli-model.fig} provides an implementation of such a
model in \proglang{Stan}.%
%
\footnote{This model is available in the \proglang{Stan} source
  distribution in \nolinkurl{src/models/basic\_estimators/bernoulli.stan}.}
%
\begin{figure}
\begin{Code}
data {
  int<lower=0> N;                  // N >= 0
  int<lower=0,upper=1> y[N];       // y[n] in { 0, 1 }
}
parameters {
  real<lower=0,upper=1> theta;     // theta in [0, 1]
}
model {
  theta ~ beta(1,1);               // prior
  y ~ bernoulli(theta);            // likelihood
}
\end{Code}
\caption{\it Model for estimating a Bernoulli parameter.}\label{bernoulli-model.fig}
\end{figure}
%
The model treats the observed binary data, \code{y[1],...,y[N]}, as
independent and identically distributed, with success probability
\code{theta}.  The vectorized likelihood statement can also be coded
using a loop as in BUGS, although it will run more slowly than the
vectorized form:
%
\begin{Code}
for (n in 1:N)
  y[n] ~ bernoulli(theta);
\end{Code}
%
A \code{beta(1,1)} (i.e., uniform) prior is placed on \code{theta},
although there is no special behavior for conjugate priors in
\proglang{Stan}.  The prior could be dropped from the model altogether
because parameters start with uniform distributions on their support,
here constrained to be between 0 and 1 in the parameter declaration
for \code{theta}.

@@

Data for running \proglang{Stan} from the command line can be included
in \proglang{R} dump format.%
%
\footnote{A \pkg{JSON} interface for structured data input and output is
  currently under development.}
%
For example, 10 observations for the
model in Figure~\ref{bernoulli-model.fig} could be encoded as%
%
\footnote{This data
file is provided with the \proglang{Stan} distrbution in file
\nolinkurl{src/models/basic\_estimators/bernoulli.R.stan}.}
%
\begin{Code}
N <- 10
y <- c(0,1,0,0,0,0,0,0,0,1)
\end{Code}
%
This defines the contents of two variables, an integer \code{N} and a
10-element integer array \code{y}.  The variable \code{N} is declared
in the data block of the program as being an integer greater than or
equal to zero;  the variable \code{y} is declared as an integer array of size
\code{N} with entries between 0 and 1 inclusive.

In \pkg{RStan} and \pkg{PyStan}, data can also be passed directly
through memory without the need to read or write to a file.

@@

\subsection{Running the sampler}

\subsubsection{Command to sample from the model}

The executable can be run with default options by specifying a path to
the data file.  The first command in the following example changes the
current directory to that containing the model, which is where the
data resides and where the executable is built.  From there, the path
to the data is just the file name \code{bernoulli.data.R}.
%
\begin{CodeChunk}
\begin{CodeInput}
> cd $stan/src/models/basic_estimators
> ./bernoulli sample data file=bernoulli.data.R
\end{CodeInput}
\end{CodeChunk}
%
For Windows, the \code{./} before the command should be removed.  This
invocation specifies the executable program, the data file, the file
to which the samples are written, the identity of the chain, and a
seed for the random number generator.

\subsubsection{Terminal output from sampler}

The output is as follows, starting with a summary of the command-line
options used, including defaults;  these are also written into the 
samples file as comments.
%
\begin{Code}
 method = sample (Default)
   sample
     num_samples = 1000 (Default)
     num_warmup = 1000 (Default)
     save_warmup = 0 (Default)
     thin = 1 (Default)
     adapt
       engaged = 1 (Default)
       gamma = 0.050000000000000003 (Default)
       delta = 0.80000000000000004 (Default)
       kappa = 0.75 (Default)
       t0 = 10 (Default)
       init_buffer = 75 (Default)
       term_buffer = 50 (Default)
       window = 25 (Default)
     algorithm = hmc (Default)
       hmc
         engine = nuts (Default)
           nuts
             max_depth = 10 (Default)
         metric = diag_e (Default)
         stepsize = 1 (Default)
         stepsize_jitter = 0 (Default)
 id = 0 (Default)
 data
   file = bernoulli.data.R
 init = 2 (Default)
 random
   seed = 4294967295 (Default)
 output
   file = output.csv (Default)
   diagnostic_file =  (Default)
   refresh = 100 (Default)

Gradient evaluation took 4e-06 seconds
1000 transitions using 10 leapfrog steps per transition would take 0.04 seconds.
Adjust your expectations accordingly!

Iteration:    1 / 2000 [  0%]  (Warmup)
Iteration:  100 / 2000 [  5%]  (Warmup)
...
Iteration: 1000 / 2000 [ 50%]  (Warmup)
Iteration: 1001 / 2000 [ 50%]  (Sampling)
...
Iteration: 2000 / 2000 [100%]  (Sampling)

  Elapsed Time: 0.00932 seconds (Warm-up)
                0.016889 seconds (Sampling)
                0.026209 seconds (Total)
\end{Code}
%
The sampler configuration parameters are echoed, here they are all
default values other than the data file.  These parameters may be set
on the command line.

\subsubsection{Help}

A description of all configuration parameters including default values
and constraints is available by executing
%
\begin{CodeChunk}
\begin{CodeInput}
> ./bernoulli help-all
\end{CodeInput}
\end{CodeChunk}
%
The sampler and its configuration are described at greater length in
the manual.

@@

The output CSV file, written by default to \code{output.csv}, starts
with a summary of the configuration parameters for the run.
%
\begin{Code}
# stan_version_major = 2
# stan_version_minor = 1
# stan_version_patch = 0
# model = bernoulli_model
# method = sample (Default)
#   sample
#     num_samples = 1000 (Default)
#     num_warmup = 1000 (Default)
#     save_warmup = 0 (Default)
#     thin = 1 (Default)
#     adapt
#       engaged = 1 (Default)
#       gamma = 0.050000000000000003 (Default)
#       delta = 0.80000000000000004 (Default)
#       kappa = 0.75 (Default)
#       t0 = 10 (Default)
#       init_buffer = 75 (Default)
#       term_buffer = 50 (Default)
#       window = 25 (Default)
#     algorithm = hmc (Default)
#       hmc
#         engine = nuts (Default)
#           nuts
#             max_depth = 10 (Default)
#         metric = diag_e (Default)
#         stepsize = 1 (Default)
#         stepsize_jitter = 0 (Default)
# id = 0 (Default)
# data
#   file = bernoulli.data.R
# init = 2 (Default)
# random
#   seed = 847896134
# output
#   file = output.csv (Default)
#   diagnostic_file =  (Default)
#   refresh = 100 (Default)
\end{Code}
%
\proglang{Stan}'s behavior is fully specified by these configuration
parameters.  By using the same version of \proglang{Stan} and these
configuration parameters, exactly the same output file can be
reproduced.  The pseudo-random numbers generated by the sampler are
fully determined by the seed (here randomly generated based on the
time of the run, with value \code{847896134}) and the identifier (here
\code{0}).  The identifier is used to advance the underlying
pseudo-random number generator a sufficient number of values that
using multiple chains with the same seed and different identifiers
will draw from different subsequences of the pseudo-random number
stream determined by the seed.

The output contiues with a CSV header naming the columns of the
output. For the default NUTS sampler in Stan 2.1.0, these are
%
\begin{Code}
lp__,accept_stat__,stepsize__,treedepth__,n_divergent__,theta
\end{Code}
%
The values for \code{lp\_\_} indicate the log probability (up to an
additive constant).  The column headed by \code{accept\_stat\_\_}
provides the Metropolis/slice acceptance statistic for each
iteration.%
%
\footnote{Acceptance is the usual notion for a Metropolis sampler such
  as HMC \citep{MetropolisEtAl:1953}.  For NUTS, the acceptance
  statistic is defined as the average acceptance probabilities of all
  possible samples in the proposed tree; NUTS itself uses a slice
  sampling algorithm for rejection \citep{Neal:2003}.}
%
The column \code{stepsize\_\_} indicates the step size
(i.e., time interval) of the simulated trajectory, while the column 
\code{treedepth\_\_} gives the tree depth for NUTS, defined as the 
log base 2 of the total number of steps in the trajectory.
The rest of the header will be the names of parameters; in this
example, \code{theta} is the only parameter.

Next, the results of adaptation are printed as comments.
%
\begin{Code}
# Adaptation terminated
# Step size = 0.783667
# Diagonal elements of inverse mass matrix:
# 0.517727
\end{Code}
%
By default, \proglang{Stan} uses the NUTS sampler with a diagonal mass
matrix.  The mass matrix is estimated, roughly speaking, by
regularizing the sample covariance of the latter half of the warmup
samplers; see \citep{Stan:2013} for full details.  A dense mass matrix
may also be estimated, or the mass matrix may be set to the unit matrix.

The rest of the file contains samples, one per line, matching the
header; here the parameter \code{theta} is the final value printed on
each line, and each line corresponds to a sample.  The warmup samples
are not included by default, but may be included with the appropriate
command-line invocation of the executable. The file ends with comments reporting the elapsed time.
%
\begin{Code}
-7.19297,1,0.783667,1,0,0.145989
-8.2236,0.927238,0.783667,1,0,0.0838792
...
-7.48489,0.738509,0.783667,0,0,0.121812
-7.40361,0.995299,0.783667,1,0,0.407478
-9.49745,0.771026,0.783667,2,0,0.0490488
-9.11119,1,0.783667,0,0,0.0572588
-7.20021,0.979883,0.783667,1,0,0.14527

#  Elapsed Time: 0.010849 seconds (Warm-up)
#                0.01873 seconds (Sampling)
#                0.029579 seconds (Total)
\end{Code}
%
It is evident from the values sampled for \code{theta} in the last
column that there is a high degree of posterior uncertainty in the
estimate of \code{theta} from the ten data points in the data file.

The log probabilities reported in the first column include not only
the model log probabilities but also the Jacobian adjustment resulting
from the transformation of the variables to unconstrained space.
Here, that is the absolute derivative of the inverse logistic
function;  see \citep{Stan:2013} for full details on all of the
transforms and their Jacobians.

\subsection{Sampler output analysis}

Before performing output analysis, we recommend generating multiple
independent chains in order to more effectively monitor convergence;
see \citep{GelmanRubin:1992} for more analysis.  Three more chains of
samples can be created as follows.
%
\begin{CodeChunk}
\begin{CodeInput}
./bernoulli sample data file=bernoulli.data.R random seed=847896134 \
            id=1 output file=output1.csv
./bernoulli sample data file=bernoulli.data.R random seed=847896134 \
            id=2 output file=output2.csv
./bernoulli sample data file=bernoulli.data.R random seed=847896134 \
            id=3 output file=output3.csv
\end{CodeInput}
\end{CodeChunk}
%
These calls illustrate how additional parameters are specified
directly on the command line following the hierarchy given in the
output.  The backslash (\code{\textbackslash}) at the end of each line indicates
that the command continues on the last line;  a caret (\code{\textasciicircum})
should be used in Windows.

The chains can be safely run in parallel under different processes;
details of parallel execution depend on the operating system and the shell or
terminal program. Note that, although the same seed is used for each chain, the
random numbers will in fact be independent as the chain identifier
is used to skip the pseudorandom number generator ahead.  See
Section~\ref{rng.section} for more information.

\proglang{Stan} supplies a command-line program \code{bin/print} to
summarize the output of one or more MCMC chains.  Given a directory
containing output from sampling, it is invoked as
%
\begin{CodeChunk}
\begin{CodeInput}
> ls output*.csv
\end{CodeInput}
\begin{CodeOutput}
output.csv	output1.csv	output2.csv	output3.csv
\end{CodeOutput}
\begin{CodeInput}
> $stan/bin/print output*.csv
\end{CodeInput}
